---
title: "ME114 Quick Background Quiz Results"
author: "Ken Benoit and Slava Mikhaylov"
output: html_document
---

**A Quick Background Quiz on Data Science**

**ME114 2015**

Circle the correct answer.  **Correct answers are in bold type.**

1.  $\hat{y}_i$ means

    1.  the error in predicting $y_i$

    2.  **predicted $y_i$**

    3.  the letter $y$ in old French

    4.  the mean of $y$

2.  Numerical precision refers to

    1.  a measure of the detail with which a quantity is expressed

    2.  a reference to the digits of a number that carry meaning, and
        can be defined in terms of the number of decimal places in a
        number

    3.  the closeness of a computer’s representation of a number to an
        actual quantity

    4.  **all of the above**

3.  JSON is

    1.  a popular programming language

    2.  the name of the killer in *Friday the 13th*

    3.  **short for Javascript Object Notation, a data format**

    4.  a API for querying databases

4.  In R, which of the following results when this statement is
    evaluated:    `mean(c(TRUE, FALSE, TRUE, TRUE))`

    1.  `TRUE`

    2.  `FALSE`

    3.  `0.5`  **(closest)**

    4.  `NA`

    **This was a "trick question" since the actual answer evaluates to 0.75.  So 0.50 was (sort of) closest to being correct.**

5.  Naive Bayes is the name of

    1.  a clustering algorithm

    2.  **a method for machine learning**

    3.  a method for fitting logistic regression models

    4.  the shy nerd you sat next to in high school science class

Your Name (Optional):
\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_


### Distribution of answers: ###

```{r}
results <- read.csv("data_science_quiz_results.csv")
summary(results)
```

A possibly easier way to view the results would be to convert the data into "long" format using the `reshape2` package, and then tabulate the answers by question using `table()`.
```{r}
require(reshape2)
results$id <- 1:nrow(results)
resultsMelted <- melt(results, id.vars = "id", variable.name= "q", value.name = "answer")
with(resultsMelted, table(q, answer, useNA = "ifany"))
```


