---
title: "Exercise Set 4 - Classification and non-linear models"
author: "Ken Benoit and Slava Mikhaylov"
output: html_document
bibliography: ../bibliography/bibliography.bib
---

@james2013 provide a set of exercises (conceptual and applied) for each chapter in their book. Applied exercises focus on linking conceptual material with practical implementation in R. Below are the applied exercises for two chapters covered in the lecture material. Part I below covers applied exercises for Chapter 4 "Classification" [@james2013, pp. 171-173] and Part II exercises for Chapter 7 "Moving Beyond Linearity" [@james2013, pp. 299-301].


Please **submit this exercise by email as an html file**  by putting your answers prior to the  **deadline of 9pm on Wednesday, August XX**. 

You will need to load the core library for the course textbook:
```{r}
library(ISLR)
```

## Part I: Classification

### Exercise 4.1

This question should be answered using the `Weekly` data set, which is part of the `ISLR` package. This data is similar in nature to the `Smarket` data from today's lab, except that it contains 1,089 weekly returns for 21 years, from the beginning of 1990 to the end of 2010.

(a) Produce some numerical and graphical summaries of the `Weekly` data. Do there appear to be any patterns?
(b) Use the full data set to perform a logistic regression with `Direction` as the response and the five lag variables plus `Volume` as predictors. Use the summary function to print the results. Do any of the predictors appear to be statistically significant? If so, which ones?
(c) Compute the confusion matrix and overall fraction of correct predictions. Explain what the confusion matrix is telling you about the types of mistakes made by logistic regression.
(d) Now fit the logistic regression model using a training data period from 1990 to 2008, with `Lag2` as the only predictor. Compute the confusion matrix and the overall fraction of correct predictions for the held out data (that is, the data from 2009 and 2010).
(e) Repeat (d) using LDA.
(f) Repeat (d) using QDA.
(g) Repeat (d) using KNN with $K=1$.
(h) Which of these methods appears to provide the best results on
this data?
(i) Experiment with different combinations of predictors, including possible transformations and interactions, for each of the methods. Report the variables, method, and associated confusion matrix that appears to provide the best results on the held out data. Note that you should also experiment with values for $K$ in the KNN classifier.

### Exercise 4.2

In this problem, you will develop a model to predict whether a given car gets high or low gas mileage based on the `Auto` data set.

(a) Create a binary variable, `mpg01`, that contains a 1 if `mpg` contains a value above its median, and a 0 if `mpg` contains a value below its median. You can compute the median using the `median()` function. Note you may find it helpful to use the `data.frame()` function to create a single data set containing both `mpg01` and the other `Auto` variables.
(b) Explore the data graphically in order to investigate the association between `mpg01` and the other features. Which of the other features seem most likely to be useful in predicting `mpg01`? Scatterplots and boxplots may be useful tools to answer this question. Describe your findings.
(c) Split the data into a training set and a test set.
(d) Perform LDA on the training data in order to predict `mpg01` using the variables that seemed most associated with `mpg01` in (b). What is the test error of the model obtained?
(e) Perform QDA on the training data in order to predict `mpg01` using the variables that seemed most associated with `mpg01` in (b). What is the test error of the model obtained?
(f) Perform logistic regression on the training data in order to predict `mpg01` using the variables that seemed most associated with `mpg01` in (b). What is the test error of the model obtained?
(g) Perform KNN on the training data, with several values of $K$, in order to predict `mpg01`. Use only the variables that seemed most associated with `mpg01` in (b). What test errors do you obtain? Which value of $K$ seems to perform the best on this data set?


### Exercise 4.3

This problem involves writing functions.

(a) Write a function, `Power()`, that prints out the result of raising 2 to the 3rd power. In other words, your function should compute $2^3$ and print out the results.

*Hint: Recall that `x^a` raises `x` to the power `a`. Use the `print()` function to output the result.*

(b) Create a new function, `Power2()`, that allows you to pass *any* two numbers, `x` and `a`, and prints out the value of `x^a`. You can do this by beginning your function with the line
```{r eval=FALSE}
Power2=function(x,a){
```
You should be able to call your function by entering, for instance,
```{r eval=FALSE}
Power2 (3,8)
```
on the command line. This should output the value of $3^8$, namely, 6,561.

(c) Using the `Power2()` function that you just wrote, compute $10^3$, $8^{17}$, and $131^3$.

(d) Now create a new function, `Power3()`, that actually *returns* the result `x^a` as an `R` object, rather than simply printing it to the screen. That is, if you store the value `x^a` in an object called `result` within your function, then you can simply `return()` this result, using the following line:
```{r eval=FALSE}
return(result)
```
The line above should be the last line in your function, before the `}` symbol.

(e) Now using the `Power3()` function, create a plot of $f(x) = x^2$. The $x$-axis should display a range of integers from 1 to 10, and the $y$-axis should display $x^2$. Label the axes appropriately, and use an appropriate title for the figure. Consider displaying either the $x$-axis, the $y$-axis, or both on the log-scale. You can do this by using `log="x"`, `log="y"`, or `log="xy"` as arguments to the `plot()` function.

(f) Create a function, `PlotPower()`, that allows you to create a plot of `x` against `x^a` for a fixed `a` and for a range of values of `x`. For instance, if you call
```{r eval=FALSE}
PlotPower (1:10,3)
```
then a plot should be created with an $x$-axis taking on values $1,2,\dots ,10$, and a $y$-axis taking on values $1^3,2^3,\dots ,10^3$.


### Exercise 4.4

Using the `Boston` data set, fit classification models in order to predict whether a given suburb has a crime rate above or below the median. Explore logistic regression, LDA, and KNN models using various subsets of the predictors. Describe your findings.


## Part II: Moving Beyond Linearity

### Exercise 4.5

In this exercise, you will further analyze the `Wage` data set coming with the `ISLR` package. 

(a) Perform polynomial regression to predict `wage` using `age`. Use cross-validation to select the optimal degree $d$ for the polynomial. What degree was chosen, and how does this compare to the results of hypothesis testing using ANOVA? Make a plot of the resulting polynomial fit to the data.
(b) Fit a step function to predict `wage` using `age`, and perform cross-validation to choose the optimal number of cuts. Make a plot of the fit obtained.


### Exercise 4.6

This question uses the variables `dis` (the weighted mean of distances to five Boston employment centers) and `nox` (nitrogen oxides concentration in parts per 10 million) from the `Boston` data. We will treat `dis` as the predictor and `nox` as the response.

(a) Use the `poly()` function to fit a cubic polynomial regression to predict `nox` using `dis`. Report the regression output, and plot the resulting data and polynomial fits.
(b) Plot the polynomial fits for a range of different polynomial degrees (say, from 1 to 10), and report the associated residual sum of squares.
(c) Perform cross-validation or another approach to select the optimal degree for the polynomial, and explain your results.
(d) Use the `bs()` function to fit a regression spline to predict `nox` using `dis`. Report the output for the fit using four degrees of freedom. How did you choose the knots? Plot the resulting fit.
(e) Now fit a regression spline for a range of degrees of freedom, and plot the resulting fits and report the resulting RSS. Describe the results obtained.
(f) Perform cross-validation or another approach in order to select the best degrees of freedom for a regression spline on this data. Describe your results.


### Exercise 4.7

This question relates to the `College` data set.

(a) Split the data into a training set and a test set. Using out-of-state tuition as the response and the other variables as the predictors, perform forward stepwise selection on the training set in order to identify a satisfactory model that uses just a subset of the predictors.
(b) Fit a GAM on the training data, using out-of-state tuition as the response and the features selected in the previous step as the predictors. Plot the results, and explain your findings.
(c) Evaluate the model obtained on the test set, and explain the results obtained.
(d) For which variables, if any, is there evidence of a non-linear relationship with the response?






## References
