---
title: "Assignment 10 - Working wth Twitter Data"
author: "Ken Benoit and Paul Nulty"
output: html_document
---

In this exercise we will step through the process of accessing data through the twitter API, and working witht the data in R.


## Exercise 10

To begin, we will try to set you up with your own twitter API key. 

(a) Follow the instructions provided in the twitter dev documentation and the walkthrough provided on the [readings](Day10prep.html) page.

The authentication for the API uses the four keys that you got after completing the application form on the twitter developers page. The function `setup_twitter_oauth()` in the R twitter api will connect your R application using these:
```{r, eval = FALSE}
library(twitteR)
setup_twitter_oauth(consumer_key = '',
                    consumer_secret ='',
                    access_token='',
                    access_secret='')
```

(b) Look at the documentation for the `twitteR` package and perform a simple search for a term.

```{r}
results <- searchTwitter('juncker', n=50)
#transform the results object into a data frame for inspection
df <- as.data.frame(t(sapply(results, as.data.frame)))
```

(c) Retrieve information about one of the users in the returned search results using the `getUser()` function. 

