---
title: "Assignment 5 - Resampling Methods"
author: "Ken Benoit and Slava Mikhaylov"
output: html_document
---

Assignments for the course focus on practical aspects of the concepts covered in the lectures. Assignments are largely based on the material covered in James et al (2013). You will start working on the assignment in the lab sessions after the lectures, but may need to finish them after class.

You will be asked to submit your assignments via Moodle by 7pm on the day of the class. We will subsequently open up solutions to the problem sets. 

You will need to load the core library for the course textbook:
```{r}
library(ISLR)
```


## Exercise 5.1: Cross-validation

Logistic regression may be used to predict the probability of `default` using `income` and `balance` on the [Default](http://bit.ly/R_ISLR_Default) dataset. Here we will estimate the test error of this logistic regression model using the validation set approach. Do not forget to set a random seed before beginning your analysis.

   (a) Fit a logistic regression model that uses `income` and `balance` to predict `default`.
   (b) Using the validation set approach, estimate the test error of this model. In order to do this, you must perform the following steps:
     i. Split the sample set into a training set and a validation set.
     
        For this you will find the `sample()` function useful.  To take a 5-observation sample of 10 items, you would index the rows using an index vector generated from `sample(10, 5)`.  For instance:
        
     
     ii. Fit a multiple logistic regression model using only the training observations.
     iii. Obtain a prediction of default status for each individual in the validation set by computing the posterior probability of default for that individual, and classifying the individual to the `default` category if the posterior probability is greater than 0.5.
     iv. Compute the validation set error, which is the fraction of the observations in the validation set that are misclassified.
   (c) Repeat the process in (b) three times, using three different splits of the observations into a training set and a validation set. Comment on the results obtained.
   (d) Now consider a logistic regression model that predicts the probability of `default` using `income`, `balance`, and a dummy variable for `student`. Estimate the test error for this model using the validation set approach. Comment on whether or not including a dummy variable for `student` leads to a reduction in the test error rate.
```{r}
testdata <- Default[1:10, ]  # select just the first 10 rows of Default
testdata
```
The create a random index variable and sample the dataframe accordingly.  `nrow()` will give you the row dimension (total number of rows) of a matrix or data.frame object.
```{r}
set.seed(100)
sampleIndex <- sample(nrow(testdata), 5)
sampleIndex

testdata[sampleIndex, ]   # just the rows randomly chosen
testdata[-sampleIndex, ]  # just the rows NOT randomly chosen
```
Use this approach to split the data into two sets, a training set, and a test set.

To compute the test error, you will predict the outcome for the test set (using the `newdata` argument in `predict.glm()`), and compare this to the actual values from the test set using `table()`.


## Exercise 5.2: Using the bootstrap

  One common use of the bootstrap estimator in statistical analysis is to estimate the sampling variance of an estimator with no known sampling distribution.  A classic example is the *median*: unlike the mean, there is no known sampling distribution for a median.  One common method for computing standard errors (or confidence intervals) from a sample for the median is to use the bootstrap.

Here is an example, using a loop:
```{r}
nreplicates <- 1000
bsresult <- numeric(nreplicates)  # set up a results variable, length 1000
for (i in 1:nreplicates) {
    bsresult[i] <- median(sample(Default$income, replace = TRUE))
}
mean(bsresult)          # the typical median from the bootstrapped replicates
median(Default$income)  # median of the actual income data, non-sampled
sd(bsresult)            # bootstrapped median of income
```

If we want to make this into a function, we could make it more general this way:
```{r}
bootstrapMedianSE <- function(variable, nreplicates) {
    bsresult <- numeric(nreplicates)  # set up a results variable, length 1000
    for (i in 1:nreplicates) {
        bsresult[i] <- median(sample(variable, replace = TRUE))
    }
    sd(bsresult)
}
```
Now we an apply this to any variable in the `Default` dataset to estimate the standard error of the median for that variable:
```{r}
bootstrapMedianSE(Default$balance, 500)
bootstrapMedianSE(Default$balance, 500)
```

### Some extra context on bootstrapping in R:

For nearly everything you can think of in R, there is probably a package for it already.  For instance:
```{r}
require(fortunes)
fortune()
```

There is a package for bootstrapping that makes the process a bit easier, but it requires that you wrap the function output to be bootstrapped in a very specific format: the first argument in the function must be `data`, and the second argument `index`.  For instance:
```{r}
# here is a function for the median
bootMedian <- function(data, index) {
  data <- data[index, ]
  median(data$income)
}
require(boot)
boot(Default, bootMedian, 1000)
```

### and now the questions to answer:

For the exercise, we consider the use of a logistic regression model to predict the probability of `default` using `income` and `balance` on the [Default](http://bit.ly/R_ISLR_Default) dataset. In particular, we will now compute estimates for the standard errors of the `income` and `balance` logistic regression coefficients in two different ways (a) using the standard formula for computing the standard errors in the [glm()](http://bit.ly/R_glm) function, and (b) using the bootstrap. Do not forget to set a random seed before beginning your analysis.

   (a) Using the [summary()](http://bit.ly/R_summary) and [glm()](http://bit.ly/R_glm) functions, determine the estimated standard errors for the coefficients associated with income and balance in a multiple logistic regression model that uses both predictors.  
   
   (b) Using the same model specification, estimate the standard errors for the coefficients of `income` and `balance` using the bootstrap. For this, you will combine the indexing functions on the dataset with the looping method shown above.  (It is up to you whether you want to code this as a general function or hard-wire the loop for this example.)  How does these differ from those computed analytically by the [glm()](http://bit.ly/R_glm) function?



## Extra Credit 1: Leave-out one cross-validation

In Sections 5.3.2 and 5.3.3 in James et al (2013), it is shown that [cv.glm()](http://bit.ly/R_cv_glm) function can be used to compute the LOOCV test error estimate. Alternatively, one could compute those quantities using just the [glm()](http://bit.ly/R_glm) and `predict.glm()` functions, and a [for()](http://bit.ly/R_Control) loop. You will now use this approach to compute the LOOCV error for a simple logistic regression model on the [Weekly](http://bit.ly/R_ISLR_Weekly) dataset. Recall how the LOOCV error is covered in the context of classification problems.

   (a) Fit a logistic regression model that predicts `Direction` using `Lag1` and `Lag2`.
   (b) Fit a logistic regression model that predicts `Direction` using `Lag1` and `Lag2` *using all but the first observation*.
   (c) Use the model from (b) to predict the direction of the first observation. You can do this by predicting that the first observation will go up if `P(Direction="Up"|Lag1, Lag2)` > 0.5. Was this observation correctly classified?
   (d) Write a for loop from $i=1$ to $i=n$ ,where $n$ is the number of observations in the data set, that performs each of the following steps:
    i. Fit a logistic regression model using all but the $i$th observation to predict `Direction` using `Lag1` and `Lag2`.
    ii. Compute the posterior probability of the market moving up for the $i$th observation.
    iii. Use the posterior probability for the $i$th observation in order to predict whether or not the market moves up.
    iv. Determine whether or not an error was made in predicting the direction for the $i$th observation. If an error was made, then indicate this as a 1, and otherwise indicate it as a 0.
   (e) Take the average of the $n$ numbers obtained in (d)iv in order to obtain the LOOCV estimate for the test error. Comment on the results.
   
   
## Extra Credit 2: Cross-validation on a simulated dataset

We will now perform cross-validation on a simulated data set.

   (a) Generate a simulated data set as follows:
```{R}
   set.seed(1)
   y=rnorm(100)
   x=rnorm(100)
   y=x-2*x^2+rnorm(100)
```
In this data set, what is $n$ and what is $p$? Write out the model used to generate the data in equation form.

   (b) Create a scatterplot of $X$ against $Y$. Comment on what you find.
   
   (c) Set a random seed, and then compute the LOOCV errors that result from fitting the following four models using least squares:
   1. $Y = \beta_0 + \beta_1 X + \epsilon$
   2. $Y = \beta_0 + \beta_1 X + \beta_2 X_2 + \epsilon$
   3. $Y = \beta_0 + \beta_1 X + \beta_2 X_2 + \beta_3 X_3 + \epsilon$
   4. $Y = \beta_0 + \beta_1 X + \beta_2 X_2 + \beta_3 X_3 + \beta_4 X_4 + \epsilon$.

Note you may find it helpful to use the `data.frame()` function
to create a single data set containing both $X$ and $Y$.

   (d) Repeat (c) using another random seed, and report your results. Are your results the same as what you got in (c)? Why?
   (e) Which of the models in (c) had the smallest LOOCV error? Is this what you expected? Explain your answer.
   (f) Comment on the statistical significance of the coefficient estimates that results from fitting each of the models in (c) using least squares. Do these results agree with the conclusions drawn based on the cross-validation results?